\documentclass[twoside]{article}
\setlength{\oddsidemargin}{0.25 in}
\setlength{\evensidemargin}{-0.25 in}
\setlength{\topmargin}{-0.6 in}
\setlength{\textwidth}{6.5 in}
\setlength{\textheight}{8.5 in}
\setlength{\headsep}{0.75 in}
\setlength{\parindent}{0 in}
\setlength{\parskip}{0.1 in}

%
% add packages if you need them here:
%

\usepackage{amsmath,amsfonts,graphicx}
\usepackage{algorithm,caption}
\usepackage[noend]{algpseudocode}
\usepackage{url}
%
% the following macro is used to generate the header.
%
\newcommand{\notes}[4]{
   \pagestyle{myheadings}
   \thispagestyle{plain}
   \newpage
   \setcounter{page}{1}
   \noindent
   \begin{center}
   \framebox{
      \vbox{\vspace{2mm}
    \hbox to 6.28in { {\bf CSC 365-07: Introduction to Database Systems
		\hfill Spring 2023} }
       \vspace{4mm}
       \hbox to 6.28in { {\Large \hfill Lecture 1:  Database and DBMS \hfill} }
       \vspace{2mm}
       \hbox to 6.28in { {\it Instructor: Alex Dekhtyar \hfill Ishaan Sathaye} }
      \vspace{2mm}}
   }
   \end{center}
}

% use these for theorems, lemmas, proofs, etc.
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newenvironment{proof}{{\bf Proof:}}{\hfill\rule{2mm}{2mm}}

\begin{document}
\notes




\section*{Introduction}
Definition of a database and DBMS in Professor Notes.

\newpage
\hfill \break
\framebox{
    \vbox{\vspace{2mm}
    \hbox to 6.28in { {\bf CSC 365-07: Introduction to Database Systems
    \hfill Spring 2023} }
    \vspace{4mm}
    \hbox to 6.28in { {\Large \hfill Lecture 2:  Relational Data Model \hfill} }
    \vspace{2mm}
    \hbox to 6.28in { {\it Instructor: Alex Dekhtyar \hfill Ishaan Sathaye} }
    \vspace{2mm}}
}

\section*{Relational Data Model}
\begin{definition}
    Relational data model is an approach to organizing collections of data
\end{definition}

\begin{itemize}
    \item Relation
    \begin{itemize}
        \item Relational Table $\longrightarrow$ \textbf{Name + Schema}
        \begin{itemize}
            \item Schema: List of attribute name + attribute type pairs
        \end{itemize}
    \end{itemize}
    \item Relational Database $\longrightarrow$ \textbf{Collection of Relations 
    tables}
    \item \textbf{Table Instance}: set of records with instantiated values of 
    the attributes
    \begin{itemize}
        \item Finite
        \item Records, rows, tuples
    \end{itemize}
\end{itemize}

One unit of data is called a \textbf{datum}.

Object, entity, event: description of one object, entity, event
\begin{itemize}
    \item \textbf{Records} consist of attributes or fields (rows in the 
    table).
    \item \textbf{Attributes} is a named container for a value of a specific type.
\end{itemize}

\subsection*{Database Table Constraint}
\begin{definition}
    Limitations of table instances
\end{definition}
\begin{itemize}
    \item \textbf{Candidate Key}: set or lists of attributes that uniquely
    define a record in a table, \textbf{minimal such set of attributes},
    made up of multiple attributes sometimes.
    \begin{itemize}
        \item \textbf{Every attribute is necessary.}
    \end{itemize}
\end{itemize}

\section*{Examples}
\subsection*{CSC 365 Example}

Course Object:
\begin{itemize}
    \item Prefix: CSC $\longrightarrow$ \textbf{String}
    \item Course \#: 365 $\longrightarrow$ \textbf{Integer}
    \item Name: Introduction to Database Systems $\longrightarrow$ \textbf{String}
    \item Description: Basic Principles, ... $\longrightarrow$ \textbf{String}
    \item Units: 4 $\longrightarrow$ \textbf{Integer}
\end{itemize}
Department Object:
\begin{itemize}
    \item Name: Computer Science and Software Engineering
    \item Abbreviation: CSSE
    \item Building: 14
    \item Room: 245
    \item College: CENG
\end{itemize}

Stringing these objects together based on relationship would 
make a \textbf{network model}.

\subsection*{Schema Example}
\begin{verbatim}
    Course(Prefix String, Course# Integer, Name String, Description
    String, Units Integer)
\end{verbatim}

\begin{table}[h]
    \centering
    \begin{tabular}{|c|c|c|c|c|}
        \hline
        Prefix & Course\# & Name & Description & Units \\
        \hline
        CSC & 365 & Introduction to Database Systems & Basic Principles, ... & 4 \\
        \hline
        CSC & 357 & Systems Programming & ... & 4 \\
        \hline
    \end{tabular}
\end{table}

\begin{verbatim}
    Department(Name, College, Building, Room): Department would also have a table as well.
\end{verbatim}

\newpage
\hfill \break 
\framebox{
    \vbox{\vspace{2mm}
    \hbox to 6.28in { {\bf CSC 365-07: Introduction to Database Systems
    \hfill Spring 2023} }
    \vspace{4mm}
    \hbox to 6.28in { {\Large \hfill Lecture 3:  RDM Cont. \hfill} }
    \vspace{2mm}
    \hbox to 6.28in { {\it Instructor: Alex Dekhtyar \hfill Ishaan Sathaye} }
    \vspace{2mm}}
}

\section*{Relational Data Model}

What makes a record unique?
\begin{itemize}
    \item \textbf{Superkey}: any set of attributes that uniquely defines a 
    record in a table
    \item \textbf{Primary Key}: candidate key chosen by you
\end{itemize}

\newpage
\hfill \break 
\framebox{
    \vbox{\vspace{2mm}
    \hbox to 6.28in { {\bf CSC 365-07: Introduction to Database Systems
    \hfill Spring 2023} }
    \vspace{4mm}
    \hbox to 6.28in { {\Large \hfill Lecture 4: SQL DDL and DML \hfill} }
    \vspace{2mm}
    \hbox to 6.28in { {\it Instructor: Alex Dekhtyar \hfill Ishaan Sathaye} }
    }
}

\section*{MySQL Access}
\begin{enumerate}
    \item Server Address = host: \textbf{mysql.labthreesixfive.com}
    \item Port: 3306
    \item username
    \item password
\end{enumerate}

MySQL Database
\begin{itemize}
    \item Namespace
    \item Collection of Tables
    \item Set of Permissions
\end{itemize}

\section*{Case Sensitivity}
\subsection*{Case Sensitive}
\begin{itemize}
    \item Table Names
    \item Database Names
\end{itemize}
\subsection*{Not Case Sensitive}
\begin{itemize}
    \item Attribute Names
    \item SQL Keywords
\end{itemize}

\section*{Types}
\begin{itemize}
    \item \textbf{Numeric Types}
    \begin{itemize}
        \item \textbf{Integer Types}
        \begin{itemize}
            \item TINYINT
            \item SMALLINT
            \item MEDIUMINT
            \item \textbf{INT}
            \item BIGINT
        \end{itemize}
        \item \textbf{Floating Point Types}
        \begin{itemize}
            \item \textbf{FLOAT}
            \item \textbf{DOUBLE(P, D)}
            \item \textbf{DECIMAL}
        \end{itemize}
    \end{itemize}
    \item \textbf{String Types}
    \begin{itemize}
        \item \textbf{Character Types}
        \begin{itemize}
            \item \textbf{CHAR(N)} $\longrightarrow$ \textbf{Fixed Length}
            \item \textbf{VARCHAR(N)} $\longrightarrow$ \textbf{Variable Length}
            \item TINYTEXT
            \item \textbf{TEXT} $\longrightarrow$ for storing large amounts of 
            text
            \item MEDIUMTEXT
            \item LONGTEXT
        \end{itemize}
    \end{itemize}
    \item \textbf{Date and Time Types}
    \begin{itemize}
        \item \textbf{Date Types}
        \begin{itemize}
            \item DATE
            \item DATETIME
            \item TIMESTAMP
            \item TIME
            \item YEAR
        \end{itemize}
    \end{itemize}
\end{itemize}

\section*{Data Definition Language (DDL)}

\textbf{Commands from DDL act upon the schema}
\begin{itemize}
    \item CREATE TABLE
    \item DROP TABLE
    \item ALTER TABLE
\end{itemize}

\subsection*{Define a Relational Table}
Aspects needed to define a table:
\begin{itemize}
    \item Table Name
    \item Attributes: Name + Type
    \item Constraints
\end{itemize}

\begin{verbatim}
    CREATE TABLE <table_name> (
        <attribute_name> <sql_type> [<single_line_constraints>],
        ...,
        <attribute_name> <sql_type> [<single_line_constraints>] [,
        <constraints>[,
        <constraints>]
    ]);
\end{verbatim}

\section*{Data Manipulation Language (DML)}

\textbf{Commands from DML act upon the instance.}
\begin{itemize}
    \item INSERT
    \item DELETE
    \item UPDATE
\end{itemize}

\subsection*{Inserting Data}
\begin{verbatim}
    INSERT INTO <table_name>(<attribute_name>, ...)
        VALUES (<value>, ...);
\end{verbatim}
Supply values in order of attribute declarations in CREATE TABLE statement.
Can omit the attribute names if values supplied are in the same order. If need 
to omit
a value then omit that attribute name as well.

\section*{More on Constraints}
\begin{itemize}
    \item \textbf{[NOT] NULL} - attribute cannot be null
    \item \textbf{UNIQUE}
    \item \textbf{PRIMARY KEY}
    \item \textbf{FOREIGN KEY}
    \item \textbf{DEFAULT $<$exp$>$} - default value for attribute
    \item \textbf{AUTO\textunderscore INCREMENT} - means that the attribute is 
    an integer and is automatically incremented
\end{itemize}

\section*{Lab 2}
MySQL Server
\begin{itemize}
    \item LabThreeSixFive.com
    \item mysql command line client
    \item IDE (DatGrip)
    \item mysql connectivity from Python
\end{itemize}

Lab 2 uses Create Table, Drop Table, and Insert.

\section*{Code from Lab}
\begin{verbatim}
show tables

CREATE TABLE Departments (
    
    DeptId INT PRIMARY KEY, 
    Abbr VARCHAR(20) UNIQUE, -- UNIQUE makes candidate key
    Name VARCHAR(128) UNIQUE,
    College CHAR(10),
    Building INT,
    Room CHAR(6),
    -- set multiple candidate keys at the bottom
    UNIQUE(Building, Room),
    -- foreign key always a separate line statement:
    -- FOREIGN KEY(College) REFERENCES colleges(abbr)
    
);

describe colleges;
SELECT * FROM colleges;

show CREATE TABLE colleges;

show CREATE TABLE Departments;

INSERT INTO Departments 
    VALUES(1, 'CSSE', 'Computer Science and Software Engineering', 'CENG', 14, '245');
    
INSERT INTO Departments(DeptId, Abbr, Name, College, Building, Room)
    VALUES(1, 'CSSE', 'Computer Science and Software Engineering', 'CENG', 14, '245');

\end{verbatim}

\newpage
\hfill \break 
\framebox{
    \vbox{\vspace{2mm}
    \hbox to 6.28in { {\bf CSC 365-07: Introduction to Database Systems
    \hfill Spring 2023} }
    \vspace{4mm}
    \hbox to 6.28in { {\Large \hfill Lecture 5: DDL and DML Continued \hfill} }
    \vspace{2mm}
    \hbox to 6.28in { {\it Instructor: Alex Dekhtyar \hfill Ishaan Sathaye} }
    }
}

\section*{DML}
\subsection*{Updating Data}
\begin{verbatim}
    UPDATE <table_name>
        SET <attribute_name> = <value>
        WHERE <condition>;
\end{verbatim}

\subsubsection*{Example}
\begin{verbatim}
    UPDATE colleges
        SET abbr = 'COSAM'
        WHERE abbr = 'COASM'
\end{verbatim}

WHERE clause is a filter that determines which rows are updated.

\subsection*{Deleting Data}
\begin{verbatim}
    DELETE FROM <table_name> -- just this is a valid command to delete all rows
        WHERE <condition>;
\end{verbatim}

\section*{DDL}
\subsection*{Altering Tables}
\begin{verbatim}
    ALTER TABLE <table_name>
        <Command> <parameters>;
\end{verbatim}
\subsubsection*{Commands}
\begin{itemize}
    \item ADD - add a column/attribute/key
    \item DROP
    \item MODIFY
    \item RENAME
\end{itemize}
\subsubsection*{Parameters}
\begin{itemize}
    \item COLUMN
    \item CONSTRAINT
    \item FOREIGN KEY
    \item PRIMARY KEY
    \item UNIQUE
\end{itemize}

Adding an attribute, dropping/adding a constraint, renaming a table, disable/enabling
keys, and modifying attributes examples are in this professor notes: \url{4-SQLDDLDML.pdf}

\newpage
\hfill \break 
\framebox{
    \vbox{\vspace{2mm}
    \hbox to 6.28in { {\bf CSC 365-07: Introduction to Database Systems
    \hfill Spring 2023} }
    \vspace{4mm}
    \hbox to 6.28in { {\Large \hfill Lecture 6: DML/DDL Cont., WHERE Clause., 
    and MySQL Conn. \hfill} }
    \vspace{2mm}
    \hbox to 6.28in { {\it Instructor: Alex Dekhtyar \hfill Ishaan Sathaye} }
    }
}

\section*{Announcements}
\subsection*{Running Scripts for Lab 2}
Can run from command line using mysql command or using mysql client. For 
running using
mysql command need to specify the database if not using default database.
\begin{verbatim}
    source script.sql
\end{verbatim}

\section*{DML/DDL}

\textbf{Data Manipulation works on instance and Data Definition works on schema.}

\subsection*{Altering a Table}
Modifying the schema. ALTER examples in class.

For CREATE TABLE, you can name constraints: 
\begin{verbatim}
    CREATE TABLE Example (
        Id int PRIMARY KEY,
        X INT,
        Y INT,
        CONSTRAINT Point UNIQUE (X, Y)
    );
\end{verbatim}

\subsection*{Updating and Deleting from Table: WHERE Clause}

Ex. Deleting in Table
\begin{verbatim}
    DELETE FROM test02
        WHERE b > c
\end{verbatim}
This deletes rows where b is greater than c.

Ex. Deleting with Scope
\begin{verbatim}
    DELETE FROM test02
        FOR EACH ROW in test01
        DO 
            DELETE FROM test02 -- delete(row, condition)
            WHERE b > c
\end{verbatim}

\subsubsection*{SQL Boolean Expressions}
\begin{itemize}
    \item 0, 1
    \item Builtin: IN(...) $\longrightarrow$ returns bool
    \item $<Expression>$ $<op1>$ $<op2>$ // can also use IN or LIKE
    \item $<Expression>$ AND $<Expression>$
    \item $<Expression>$ OR $<Expression>$
    \item NOT $<Expression>$
\end{itemize}


\section*{MySQL Connectivity}
Breifly went over the Python examples on Course webpage that connect to 
MySQL server.

\newpage
\hfill \break 
\framebox{
    \vbox{\vspace{2mm}
    \hbox to 6.28in { {\bf CSC 365-07: Introduction to Database Systems
    \hfill Spring 2023} }
    \vspace{4mm}
    \hbox to 6.28in { {\Large \hfill Lecture 7: Python Connectivity and 
    Relational Algebra \hfill} }
    \vspace{2mm}
    \hbox to 6.28in { {\it Instructor: Alex Dekhtyar \hfill Ishaan Sathaye} }
    }
}

\section*{Python MySQL Connectivity}
Relational Database is sitting on a server. It is listening
for connections, and our program is a client that connects
to the server via the port. Essentially, there is a pipe and a
exchange of messages that is happening. Generates a connection object
that stores info about how to properly access the database.

\subsection*{Package}
\begin{verbatim}
    import mysql.connector
\end{verbatim}

\subsection*{Connection}
5 Things Needed: Host, Port, Username, Password, Database (sometimes not 
necessarily)

These get passed to mysql.connector.connect() function. This returns a 
connection object.
is\textunderscore connected() returns a cursor object.

Cursor object that is returned from the connection object. Cursor object is 
used to execute queries.


\section*{Relational Algebra}
Relational $\longrightarrow$ Database Model $\longrightarrow$ Relational Model.
Algebra: set of elements \& operations on elements

Relational Algebra is operations on relational tables.

Boolean Algebra introduces operations on truth values
\begin{itemize}
    \item T, F
    \item \textasciitilde, $\wedge$, $\vee$, $\rightarrow$, $\leftrightarrow$
\end{itemize}

\subsection*{Notation}
Upper case letters like R, S, T, $R_1$, $S_7$, ... are relational table names.
Letters from first half of alphabet like A, B, C, ... are attributes names.
R($A_1$,..., $A_n$) are to represent schema. t, s, r $\in$ R are tuples. $a_1$, $a_2$, ...
are values. Ex. t = ($a_1$,..., $a_n$) and it could be referred to as t.$A_1$ = $a_1$.

\subsection*{Operations}
\subsubsection*{Binary}
\subsubsection*{Unary}
\begin{itemize}
    \item Selection $\sigma$ - filter rows
    \item Projection $\pi$ - filter columns
\end{itemize}

\subsection*{Selection Operation}
\begin{itemize}
    \item $\sigma_{<selection condition>}(R)$ - returns rows that satisfy condition
    \item Selection Condition denoted by $C$
    \item Ex. C = $A_2 = 'Riley' \wedge A_3 = 'Hicks'$
    \item \textbf{Formal Notation:} $\sigma_C(R)$ = $\{t \in R | t  satisfies C\}$
\end{itemize}

\subsection*{Projection Operation}
\begin{itemize}
    \item $\pi_{<attribute list>}(R)$ - returns columns that are in attribute list
    \item $F$ is the projection list which is a list of attributes
    \item Ex. F = $(B_1,..., B_m)$ where $B_i$ $\in$ $A_1,..., A_n$
    \item \textbf{Formal Notation:} $\pi_F(R)$ = $\{t' | \exists,  t \in R, s.t. 
    \forall B \in F, t'.B=t.B\}$
    \item \textbf{Projection squeezes out duplicates}
\end{itemize}

\newpage
\hfill \break 
\framebox{
    \vbox{\vspace{2mm}
    \hbox to 6.28in { {\bf CSC 365-07: Introduction to Database Systems
    \hfill Spring 2023} }
    \vspace{4mm}
    \hbox to 6.28in { {\Large \hfill Lecture 8: Relational Algebra \hfill} }
    \vspace{2mm}
    \hbox to 6.28in { {\it Instructor: Alex Dekhtyar \hfill Ishaan Sathaye} }
    }
}

\section*{Relational Algebra}
R, S,  T are relational tables and these are sets. Set Operations include
Union, Intersection, Difference, Symmetric Difference...

Example:

\begin{table}[!htb]
    \begin{minipage}{.5\linewidth}
      \caption{R}
      \centering
        \begin{tabular}{c|c|c}
            A & B & C \\
            \hline
            1 & 2 & a \\
            2 & 4 & b \\
            3 & 1 & d \\
            4 & 4 & d \\
            4 & 5 & a \\
        \end{tabular}
    \end{minipage}
    \begin{minipage}{.25\linewidth}
      \centering
        \caption{S}
        \begin{tabular}{c|c|c}
            B & D & E \\
            \hline
            2 & a & 1 \\
            3 & a & 2 \\
            7 & b & 3 \\
            5 & b & 1 \\
            1 & c & 2 \\
        \end{tabular}
    \end{minipage} 
\end{table}

\subsection*{Selection Review}
Duplicate Elimination: $\sigma_{A=4 \vee B=4}(R)$

\begin{table}[!htb]
    \centering
    \begin{tabular}{c|c|c}
        A & B & C \\
        \hline
        2 & 4 & b \\
        4 & 4 & d \\
        4 & 5 & a \\
    \end{tabular}
\end{table}

\subsection*{Projection Review}
$\bullet$ Subset and Columns: $\pi_{B,D}(R)$. The schema here is $R(B,D)$.

$\bullet$ Can make composition of operations: $\pi_{B,E}(\sigma_{D=b}(R))$ because 
an result of an operation is a relational table. Result:
\begin{table}[!htb]
    \centering
    \begin{tabular}{c|c}
        B & E \\
        \hline
        7 & 3 \\
        5 & 1 \\
    \end{tabular}
\end{table}

$\bullet$ Can also reorder columns: $\pi_{EBD}(R)$

$\bullet$ Duplicate Columns: $\pi_{ABA}(R)$. Issue with this is that there are 2
columns with the same name. Disambiguate by renaming: $\pi_{A_1BA_2}(R)$

$\bullet$ Can also introduce new columns: $\pi_{A,B,2\cdot A}(R)$

\subsection*{Cartesian Product}
$\bullet$ $R \times S$ is the cartesian product of R and S. Result is a table 
with all possible combinations of rows from R and S. 

$\bullet$ Notation: $R \times S$ = $\{(t,t`) | t \in R, t \in S\}$

Ex. $\sigma_{A<3}(R) \times \sigma_{B<3}(S)$
\begin{table}[!htb]
    \centering
    \begin{tabular}{c|c|c|c|c|c}
        A & B & C & B & D & E \\
        \hline
        1 & 2 & a & 2 & a & 1 \\
        1 & 2 & a & 3 & a & 2 \\
        1 & 2 & a & 5 & b & 1 \\
        2 & 4 & b & 2 & a & 1 \\
        2 & 4 & b & 3 & a & 2 \\
        2 & 4 & b & 5 & b & 1 \\
    \end{tabular}
\end{table}

\subsection*{Join}

\begin{table}[!htb]
    \begin{minipage}{.5\linewidth}
      \caption{R}
      \centering
        \begin{tabular}{c|c|c}
            Id & Customer & C \\
            \hline
            1 & 2 & a \\
            2 & 4 & b \\
            3 & 1 & d \\
            4 & 4 & d \\
            4 & 5 & a \\
        \end{tabular}
    \end{minipage}
    \begin{minipage}{.25\linewidth}
      \centering
        \caption{S}
        \begin{tabular}{c|c|c}
            Id & Name & E \\
            \hline
            2 & a & 1 \\
            3 & a & 2 \\
            7 & b & 3 \\
            5 & b & 1 \\
            1 & c & 2 \\
        \end{tabular}
    \end{minipage} 
\end{table}

$\bullet$ Who purchased Receipt 3 $\longrightarrow$ Find a person with Id 1: $
\pi_{Name}(\pi_{R.Cust=S.Id}(\sigma_{Id=3}(R))\times S)$

$\bullet$ Notation: $R \bowtie_{C} S = \sigma_{R.x operation S.Y}(R \times S)$

\newpage
\hfill \break
\framebox{
    \vbox{\vspace{2mm}
    \hbox to 6.28in { {\bf CSC 365-07: Introduction to Database Systems
    \hfill Spring 2023} }
    \vspace{4mm}
    \hbox to 6.28in { {\Large \hfill Lecture 9:  Joins \hfill} }
    \vspace{2mm}
    \hbox to 6.28in { {\it Instructor: Alex Dekhtyar \hfill Ishaan Sathaye} }
    \vspace{2mm}}
}

\section*{Announcements}
$\bullet$ Double-sided cheat sheet allowed for Midterm.
$\bullet$ Homework assigned on Thursday as prep for Midterm.

\section*{Example}

\begin{table}[!htb]
    \begin{minipage}{.5\linewidth}
      \caption{T}
      \centering
        \begin{tabular}{c|c|c|c}
            Id & Name & Course & Grade \\
            \hline
            1 & Joe & CSC 365 & A \\
            2 & Mary & CSC 365 & A \\
            3 & Lisa & CSC 101 & B \\
            4 & Luis & STAT 301 & C \\
            5 & Steve & STAT 301 & A \\
            1 & Joe & CSC 357 & B \\
            2 & Mary & CSC 357 & C \\
        \end{tabular}
    \end{minipage}
    \begin{minipage}{.25\linewidth}
      \centering
        \caption{C}
        \begin{tabular}{c|c}
            CourseId & Instructor \\
            \hline
            CSC 357 & Nico \\
            CSC 365 & Migler \\
            STAT 301 & Bodwin \\
            CSC 101 & Rivera \\
            STAT 302 & Frame \\
        \end{tabular}
    \end{minipage} 
\end{table}

Transcript Table T has primary key $(Id, Course)$ and foreign key $(Course)$ 
referencing Course Table C.

$\bullet$ \textbf{Find all Joe's instructors.}
\begin{gather*}
    \sigma_{Name=Joe}(T) \\
    \pi_{Course}(\sigma_{Name=Joe}(T)) \\
    C \times \pi_{Course}(\sigma_{Name=Joe}(T)) \\
    \sigma_{C.CourseNo=T.Course}(C \times \pi_{Course}(\sigma_{Name=Joe}(T))) \\
    \pi_{Instructor}(\sigma_{C.CourseNo=T.Course}(C 
    \times \pi_{Course}(\sigma_{Name=Joe}(T)))) \\
\end{gather*}

Cartesian Product in line 3 makes table with Course No., Instructor, and Course
combination for Joe. So it would be every combination with Course and CourseNo.:
\begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c}
        C.CourseNo & C.Instructor & T.Course \\
        \hline
        CSC 357 & Nico & CSC 365 \\
        CSC 365 & Migler & CSC 365 \\
        ... & ... & ... \\
        CSC 357 & Nico & CSC 357 \\
        CSC 365 & Migler & CSC 357 \\
        ... & ... & ... \\
    \end{tabular}
\end{table}

In Line 4, we are selecting the rows where the CourseNo. in C is equal to the
Courses in T. 

$\bullet$ \textbf{For each course grade, report the instructor who assigned it.}
\begin{gather*}
    \pi_{T.Course, T.Grade, C.Instructor}(\sigma_{T.Course=C.CourseId}(T 
    \times C))
\end{gather*}
First, we are taking the Cartesian product which gives us 35 rows. Then we are
making the table smaller by only selecting the rows that are equal in Courses.

\section*{$\Theta-$Join}
\begin{gather*}
    R \bowtie_{\Theta} S = \sigma_{\Theta}(R \times S) \\
\end{gather*}
\begin{center}
    $\Theta$ - selection condition where each comparison uses attributes from 
    R and S
\end{center}

\section*{Equi-Join}
\begin{gather*}
    R \bowtie_{\Theta} S \\
\end{gather*}
\begin{center}
    $\Theta$ = (R.A = S.B) $\wedge$ (R.C = S.D) $\wedge$ (R.E = S.F) \\
\end{center}

\section*{Natural Join}
\begin{center}
    $R(A_1,...A_k, B_1,...B_k)$ \\
    $S(B_1,...B_k, C_1,...C_k)$ \\
\end{center}
\begin{gather*}
    R \bowtie S = \pi_{R.*, S.C_1,...S.C_k}(R \bowtie_{Condition} S)
\end{gather*}
\begin{center}
    Condition is where $R.B_1$ = $S.B_1$ $\wedge$ ... $\wedge$ $R.B_k$ = $S.B_k$.
\end{center}
Natural join looks at all \textbf{common attributes} of two relations and joins on
them, removing one set of common attributes from the final relation.

\section*{Semi-Join}
Special case of natural join where only attributes of one relation are kept.
Essentially, it is a projection on all elements of 1 relation.

Left Semi Join:
\begin{gather*}
    R \bowtie_{\Theta} S = \pi_{R.*}(R \bowtie_{\Theta} S) \\
\end{gather*}

Right Semi Join:
\begin{gather*}
    R \bowtie_{\Theta} S = \pi_{S.*}(R \bowtie_{\Theta} S) \\
\end{gather*}

Example:
\begin{gather*}
    (\sigma_{Name=Joe}(R) \bowtie C) \\
\end{gather*}


\end{document}
